Some examples in extended logic counting classes

### 2016-09-14 QSO/logics.tex
14 sep

We introduce the operator $[\textbf{alfp}_{(x_1,\ldots,x_k),R}\varphi(x_1,\ldots,x_k,R)\mid\alpha(u_1,\ldots,u_k,R,p)](y_1,\ldots,y_k)$ which uses a FO-formula $\varphi$ as well as a QSO-formula $\alpha$ and takes as input a placeholder variable $p$, which is interpreted as a constant. The function $\zeta_{\varphi,\alpha}:2^{A^k} \to \nat$ is defined as $\zeta_{\varphi,\alpha}(T_{\varphi}(X)) = \sem{\alpha'}(\A,R)$ where $\alpha'$ is obtained by replacing every instance of $p$ by $\zeta_{\varphi,\alpha}(X)$.

>>>>

We introduce the operator $[\textbf{alfp}_{\varphi(x_1,\ldots,x_k,R)}\alpha(u_1,\ldots,u_k,R,\pi)](y_1,\ldots,y_k)$ which uses a FO-formula $\varphi$ as well as a QSO-formula $\alpha$ and takes as input a placeholder function $\pi$, which is interpreted as a constant. The partial function $\zeta_{\varphi,\alpha}:2^{A^k} \to \nat$ is such that  $\zeta_{\varphi,\alpha}(T_{\varphi}(X)) = \sem{\alpha'}(\A,R)$ where $\alpha'$ is obtained by replacing every instance of $\pi(\bar{z})$ by $\zeta_{\varphi,\alpha}(X)$, and $\zeta_{\varphi,\alpha}(\emptyset) = 0$. Then the operator is evaluated as $$\sem{[\textbf{alfp}_{\varphi(x_1,\ldots,x_k,R)}\alpha(u_1,\ldots,u_k,R,\pi)](y_1,\ldots,y_k)}(y_1,\ldots,y_k) = \zeta_{\varphi,\alpha}(X^*)$$, where $X^*$ is the least fixed point of $T_{\varphi}$.

As an example, we will show a formula that counts the number of paths of size $n$ of a structure $\A$ with a binary relation $E$. First we define $\varphi(x,R)$:
$$
\varphi(x,R) = \forall y(x < y \vee x = y) \vee \exists z(R(z) \wedge \varphi_{succ}(z,x)),x
$$
where $\varphi_{succ}(x,y)$ is a formula that is satisfied by pairs $(x,y)$ that are consecutive in the order $<$. That is, $\varphi_{succ}(x,y) = x < y \wedge \forall z((x < z \wedge z < y) \to (x = z \vee z = y) )$. Now we define $\alpha(R)$ as:
$$
\alpha(x,y,R) = \neg \exists z(R(z)) + \Sigma z[E(x,z)\cdot \pi(z,y)].
$$
Then, the formula $[\alfp_{\varphi(x,R)} \alpha(x,y,R)](x,y)$ when evaluated on $(\A,a,b)$ will count the number of paths of size $n$ from $a$ to $b$.

We also define the less powerful operator $\pth$ as follows. Given a positive formula $\varphi(x,R)$ and a formula $\psi(x,y)$ we define $[\pth_{\varphi(x,R)} \psi(x,y)](x,y)$ as a formula that when evaluated on $(\A, a, b)$ induces a graph over $\A$ with the binary relation determined by $\psi(x,y)$ and counts the number of paths from $a$ to $b$ of a size determined by the $\varphi(x,R)$. Formally, we define:
$$
\sem{[\pth_{\varphi(x,R)} \psi(x,y)](x,y)} = \sem{[\alfp_{\varphi(x,R)} \alpha_{\psi}(x,y,R)](x,y)},
$$
where $\alpha_{\psi}(x,y,R) = \neg \exists z(R(z)) + \Sigma z[\psi(x,z)\cdot \pi(z,y)]$.

\begin{theorem}
	Given a positive $\fo$ formula $\varphi(\bar{x},R)$ and a $\qfo$ formula $\alpha(\bar{x})$, there exists a $\qso$ formula $\beta(\bar{x})$ such that $\sem{[\alfp_{\varphi(\bar{x},R)} \alpha_{\psi}(\bar{x},R)](\bar{x})} = \sem{\beta(\bar{x})}$.
\end{theorem}
\begin{proof}

\end{proof}
